\plaintitle

### 2016-09-20 Paper/figures/figures.tex
Updates to general methods.

\newcommand{\outlierFig}{
 \begin{figure}
 \centering
 \includegraphics[width=.55\columnwidth]{figures/exp4}
 \caption{
 	A cropped example stimulus from experiment 3. We replaced the final 15 points of this dataset with extreme values. The overlaid purple trend line represents a robust OLS fit (ignoring the outlier values), while the overlaid green line represents the fit with all points included. Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.
 }
 \label{fig:outlier}
 \end{figure}
}

\newcommand{\expFig}{
	\begin{figure}
		\centering
		\includegraphics[width=.65\columnwidth]{figures/exp}
		\caption{
			An example estimation task from our experiments. Here, the participant must adjust the amplitude of the purple trigonometric function until it best matches a particular set of bivariate data.
		}
		\label{fig:experiment}
	\end{figure}
}

>>>>

\newcommand{\outlierFig}{
  \begin{figure}
  \centering
  \includegraphics[width=.55\columnwidth]{figures/exp4}
  \caption{A cropped example stimulus from experiment 3. We replaced the final 15 points of this dataset with extreme values. The overlaid purple trend line represents a robust OLS fit (ignoring the outlier values), while the overlaid green line represents the fit with all points included. Participants' estimates of trend lines were universally closer to the robust than the non-robust trend; regression by eye tends to underweight outliers compared to OLS.}
  \label{fig:outlier}
  \end{figure}
}

\newcommand{\expFig}{
  \begin{figure}
  \centering
  \includegraphics[width=.65\columnwidth]{figures/exp}
  \caption{An example estimation task from our experiments. Here, the participant must adjust the amplitude of the purple trigonometric function until it best matches a particular set of bivariate data.}
  \label{fig:experiment}
  \end{figure}
}

### 2016-09-20 Paper/proceedings.tex
Updates to general methods.

The estimation of trends in bivariate data is an important analytical task, as it is the basis for many factors relevant to decision-making, such as prediction, imputation, and comparison. However, model information is not always explicitly included by visualization designers. When designers do include trend information (for instance, by annotating a scatterplot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent. Viewers must therefore perform visual estimation to gain a sense of any relevant statistics not provided.

@@

In order to assess the ability of viewers of visualizations to estimate trend, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform. We designed these experiments to establish a baseline for performance at regression by eye, and to examine potential sources of bias for these estimates. Crowd-sourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}. Crowd-sourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{talbot2014four}.

In each experiment, we presented participants with a set of bivariate visualizations based on an underlying set of points. We wished to have precise and orthogonal control over relevant statistics such as noise, slope, and sample size, while maintaining the appearance of a ``natural'' distribution of points. Existing methods for generating points for related experimental tasks such as estimation of correlations (e.g., Harrison et al.~\cite{harrison2014ranking}) do not afford this orthogonality: e.g., Pearson's $r$ is correlated with the slope (points with identical residuals but different slopes of their linear fits will have differing $r$ values). We also wished to have a fair comparison between visual estimations and the results of OLS. We therefore used the standard model of OLS to generate points, namely that $y=\beta x + \epsilon$, where $\epsilon$ is an error term that follows a Gaussian model.

Therefore, for each point set, we created a set of residual values, sampled evenly from a discretized Gaussian. This set was permuted, and then applied to the points along with a target linear trend. As heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$. In all experiments, we desired control over the direction of the trend. For linear fits, this is the slope of the trend line. For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term. For trignomoetric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes). These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.

In all experiments, participants entered in information using a slider without tick marks, in order to avoid anchoring effects \cite{matejka2016effect}. Moving the slider adjusts a purple trend line by adjusting its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The central measure is accuracy, or the difference between the selected slope and the slope of the series if the residuals were removed. We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model-space of participants, and to reduce the impact of factors such as accuracy in motor movements and heterogeneity in input devices.

@@

We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For validation, certain trials contained the actual (OLS) trend line. We excluded the results of participants with poor performance on these validation questions. Across the three reported experiments and pilots, we performed 7 such exclusions. We recruited additional participants to replace these poor performers. Based on timings from internal piloting, we paid each participant \$2 for their participation, for an intended rate of \$8/hour.

We analyzed 48 participants for each experiment (excluding rejections), for a total of 144 (98 male, 43 female, 3 who declined to stat, $M_{age}= 33.2$,$SD_{age}=8.8$) participants. The 10 had graduate or professional degrees, 69 had college degrees, 40 had at least some college, and the remaining 25 had high school diplomas. After completing the main experimental task, participants were asked to self-assess their familiarity with graphs and charts, on a 5 point Likert scale. The plurality (64) rated themselves as ``3. Some familiarity,'' and none rated themselves with the maximum rating of ``5. A great deal of experience.''

@@

We presented participants with a series of bivariate visualizations, and adjusted a slider to fit the perceived trend in the points. Visualizations were one of a scatterplot, linegraph, or area chart (with the filled area below the line). Visualizations consisted of one of three types of trend (linear, quadratic, or trigonometric). For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).

Participants saw one of each combination of 3 chart types (scatterplot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli, with an additional 4 validation stimuli otherwise excluded from analysis. Type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatterplots.
\end{enumerate}
\subsection{Results}
We performed a one-way ANCOVA of the effect of graph type on \emph{signed} error. We included residual bandwidth as a covariate, and participant ID as a random factor.

Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatterplots or linegraphs}. Graph type was a significant main effect on signed error ($F(2,2431)=27$, $p<0.001$). A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatterplots and linegraphs. The trimmed mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions. This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for linegraphs and scatterplots). Figure \ref{fig:outlier} illustrates this result.

@@

In this paper, we examine the ability of viewers of visualizations to perform regression by eye --- the visual estimation of trend in bivariate visualizations. We show that even viewers without statistical training can reliably estimate even non-linear trends in visualizations like scatterplots and linegraphs. However, area charts are subject to within the bar bias, and promote trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than simple statistical regression techniques, resulting in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, can give equal weight all points in a dataset.

>>>>

The estimation of trends in bivariate data is an important analytical task, as it is the basis for many factors relevant to decision-making, such as prediction, imputation, and comparison. However, model information is not always explicitly included by visualization designers. When designers do include trend information (for instance, by annotating a scatter plot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent. Viewers must therefore perform visual estimation to gain a sense of any relevant statistics not provided.

@@

In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk platform. We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias for these estimates. Crowdsourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}. Crowdsourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}. In this section, we describe experimental design aspects shared across all experiments.

@@

We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For validation, certain trials contained the actual (OLS) trend line. We excluded the results of participants with poor performance on these validation questions. Across the three reported experiments and pilots, we performed 7 such exclusions. We recruited additional participants to replace these excluded subjects. Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.

We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $\mu_{age}= 33.2$, $\sigma_{age}=8.8$). Across the subject pool, 10 reported having graduate or professional degrees, 69 college degrees, 40 at least some college, and 25 high school diplomas. After completing the main experimental task, participants were asked to self-assess their familiarity with charts and graphs on a 5 point Likert scale. The plurality (64) rated themselves as ``3. Some familiarity,'' and none rated themselves with the maximum rating of ``5. A great deal of experience.''

\subsection{Experimental Interface}

In each experiment, we presented participants with visualizations of bivariate data and asked them to interactively adjust trend lines to best fit that data. Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model space and to reduce the impact of factors such as accuracy of motor movements or heterogeneity of input devices.

\subsection{Data Generation}

For each stimulus we wished to have precise, independent control over relevant statistics such as noise, slope, and sample size, while maintaining the appearance of a ``natural'' distribution of points. Existing methods for generating points for related experimental tasks such as estimation of correlations (e.g., Harrison et al.~\cite{harrison2014ranking}) do not afford independence: e.g., Pearson's $r$ is correlated with the slope, as points with identical residuals but different slopes of their linear fits will have differing $r$ values. We also wished to have a fair comparison between visual estimation and the results of ordinary least squares (OLS) regression. We therefore used the standard model of OLS to generate points, namely that $y=\beta x + \epsilon$, where $\epsilon$ is a normally-distributed error term.

For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian. This set was permuted, and then applied to the points along with a target trend. As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$. In all experiments, we desired control over the direction of the trend. For linear fits, this is the slope of the trend line. For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term. For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes). These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.

@@

We presented participants with a series of bivariate visualizations, and adjusted a slider to fit the perceived trend in the points. Visualizations were one of a scatter plot, linegraph, or area chart (with the filled area below the line). Visualizations consisted of one of three types of trend (linear, quadratic, or trigonometric). For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).

Participants saw one of each combination of 3 chart types (scatter plot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli, with an additional 4 validation stimuli otherwise excluded from analysis. Type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatter plots.
\end{enumerate}
\subsection{Results}
We performed a one-way ANCOVA of the effect of graph type on \emph{signed} error. We included residual bandwidth as a covariate, and participant ID as a random factor.

Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatter plots or linegraphs}. Graph type was a significant main effect on signed error ($F(2,2431)=27$, $p<0.001$). A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and linegraphs. The trimmed mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions. This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for linegraphs and scatter plots). Figure \ref{fig:outlier} illustrates this result.

@@

In this paper, we examine the ability of viewers of visualizations to perform regression by eye --- the visual estimation of trend in bivariate visualizations. We show that even viewers without statistical training can reliably estimate even non-linear trends in visualizations like scatter plots and linegraphs. However, area charts are subject to within the bar bias, and promote trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than simple statistical regression techniques, resulting in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, can give equal weight all points in a dataset.
